---
title: "RIR: Southeast Alaska red king crab mark recapture experiments"
author: "Katie Palof"
date: "March 19, 2018"
output: html_document
---

# INTRODUCTION

Current survey methods and results
Red king crab biology / management 


# OBJECTIVES
1) Estimate the population size of red king crab survey areas using mark-recapture methods
2) Compare the estimates of population size from the mark-recapture experiments to those obtained from the annual red king crab survey using a CSA model.
3) Calculate an adjustment between population estimates from the catch-survey analysis (CSA) model and the mark-recapture experiment

# METHODS

## Data Analysis

### Mark-Recapture Estimation
Mark/recapture estimate are a well established way of determining population size. The Chapman estimator (1951) was used to estimate abundance in this study since it is less bias than the Petersen estimator.  However the Chapman estimator is unbiased only if $n_1+n_2>N$.

$\hat{N}=\frac{(n_1+1)(n_2+1)}{(m_2+1)}-1,$

where

$n_1$ = the number of tagged crabs released,  
$n_2$ = the number of crabs examined at the second event,  
$m_2$ = the number of crabs with tags present at the second event, and,  
$\hat{N}$ = the estimated abundance.

There are a suite of assumptions that are included in a mark-recapture study:

1.  The population is closed (no mortality, immigration, emigration, recruitment, etc.)
2.  All individuals have the same probability of capture $p_1$ at time $t_1$
3.  The sample of time $t_2$ is a simple random sample without replacement
4.  Marking doesn’t affect probability of capture $p_2$
5.  There is no tag loss
6.  All individual with marks are reported


For these estimators, when the expected number of recaptures *is less than 10* the estimate of abundance may fail to give even the order of magnitude of the population correctly (Seber 1982).  Also, estimates of abundance (N) with low recaptures ($m_2$) have very large variances and therefore do not have much confidence associated with them.

When there have been multiple tagging and recapture periods different abundance estimators can be used. The advantage of using these would be to reduce the variance of the abundance estimate. In theory, the lower the variance the better the estimator. It is rare that such estimators change the point estimate much from the Chapman estimate, but they have better overall properties. 

Two common multi-stage abundance estimators are the pooled Petersen, that is the unweighted mean of Chapman estimates for each time period, or the Schnabel estimator. The Schnabel estimator is a mean of each time period weighted by the number of recaptures. The Schnabel may reduce the variance of a population estimate whereas the pooled Petersen cannot when compared to a single Chapman estimate. 

Schnabel estimator:


$\hat{N}=\frac{lamda}{m+1},$
where

$lamda$ = $\sum_{i=1}^s n_iM_i$  
$m$ = the total number of individuals marked,  
$M$ = the number of marked individuals at time period *i*,   
$n_i$ = the number of individuals captured in time period *i*, and,   
$s$ = the number of time periods.

### Adjustment Between Mark Recapture and CSA Estimates
An objective of this study was to compare the estimates of population size from the mark recapture experiments to those obtained from the annual red king crab survey using a catch-survey analysis model(CSA). One method to do so is to establish an adjustment factor between these two estimates. The two population estimates have to be in the same units to compare them.  The mark-recapture estimates legal crab in numbers of individuals and the CSA estimates legal crab in lbs (using the average weight of a legal crab). Therefore, the average weight of legal crab from the survey was used to convert the number of legal crab into a biomass (lb, Table 1).  The average weight of an individual legal crab is determined by applying an length-weight relationship to all legal crab caught in the survey and taking an average of the resulting weights, each survey area has its own length-weight relationship (reference with red crab ROP or Clark et al. 2003).  


**Table 1: Average weight of individual legal crab from survey data**

```{r setup, echo=F, message = F}
library(tidyverse)
library(pander)
library(knitr)
library(data.table)
# remove index column from this table *FLAG*
data <- read.csv("../data/mr_rinput.csv", header = TRUE)  
data %>% arrange(year) -> data
# M = number of animals marked on first visit
# n = number of animals captured on the second visit
# m = number of recaptured animals (marked recaped on second visit)
data %>% dplyr::select(year, area, legal_wt) ->weight
weight2 <- weight[c(1:4,6:10,12:13), ] # remove holkham no CSA estimate, and only one line for seymour 2015
pander(weight2, row.names =F, digits =3)
#kable(weight2, digits = 2, row.names = F)
```

These weights allow for both the mark-recapture and the CSA estimates to be reflected in biomass (lb) of legal crab.  Once these population estimates are on the same scale an adjustment factor was calculated by dividing the mark-recapture population estimate in pounds ($\hat{B}$) by the CSA estimate of abundance ($B$). 

$Adj=\frac{\hat{B}}{B}$

Ideally, if this adjustment factor hold for all population levels and conditions then it could be used to adjust future CSA estimates to the level of the mark-recapture if needed.  Applying this adjustment over multiple years carries a few assumptions: 

1. The mark-recapture estimate of population size is estimated without bias or error.
2. The CSA estimate of population size is estimated without bias or error
3. The relationship between the two estimates will be similar in periods of small and large population size. 



# RESULTS
Mark recapture experiments were conducted over a number of years (2009-2017) with the goal of replicating the experiment at least twice in each area.  The results of the experiments, along with their associated estimates of numbers of legal crabs (Chapman or Schnabel) are in Tables 2a and 2b. Lower and upper confidence intervals for the mark recapture estimates are provided also. 

One of the goals of this project was to compare the mark recapture experiments to the biomass estimates obtained from the CSA model, which uses our pot survey data. The CSA outputs legal biomass (in lbs) while the mark recapture (MR) outputs the estimates as number of crab.  The number of crab are then converted to biomass using the average weight of legal crab that year in our survey (Table 1). Table 3 summarizes both the estimates (in lbs) and shows the lower (LCI) and upper (UCI) confidence intervals for the mark recapture estimates.

An additional goal was to provide an adjustment to our current CSA estimate based on the differences between the CSA and the mark recapture estimates. Table 4 lists this adjustment and the associated lower and upper confidence intervals, which are based on the error associated only with the mark recapture estimate of biomass at this time. Currently, the stock health document adjusts the CSA or survey biomass using a weighted average of these adjustment calculations. These are listed in Table 5. 


**Table 2a: Summary of mark recapture experiment and Chapman population estimates (in numbers of legal crab)**
```{r data, echo=F, message = F}
data %>% 
  mutate(Chapman = (((M+1)*(n+1))/ (m+1))-1, 
         SE = (((M+1)*(n+1)*(M-m)*(n-m))/((m+1)*(m+1)*(m+2)))^0.5, 
         upper = Chapman + (SE*1.96), lower = Chapman -(SE*1.96)) ->data_sum
data_sum %>% 
  mutate(Chap_lb = Chapman*legal_wt, lower_lb = lower*legal_wt, upper_lb = upper*legal_wt, 
         adj = Chap_lb/ CSA_legalcrab) -> data_sum2

data_sum2[c(1:4, 6:9, 12:13), ] -> data_sum_Chaponly
data_sum2[c(1:9, 12:13), ] -> data_sum_Chaponly2 # only remove Seymour 2015, keep Holkham here but remove in next tables due to lack of CSA estimate
data_sum_Chaponly2 %>% 
  dplyr::select(year, area, M, n, m, Chapman, lower, upper) ->table2

panderOptions("digits", 4)
pander(table2, row.names = F )
#df <- data.frame(Year=table2$year,Area = table1$area, marked_n1 = table1$M, n2 = table1$n, 
#                 recap_m2 = table1$m, N_Chapman = table1$Chapman, lowerCI = table1$lower, 
#                 upperCI = table1$upper)
```

**Table 2b: Summary of mark recapture experiment and Schnabel population estimates (in numbers of legal crab)**
```{r schnabel, echo=F, message = F}
#Schabel Estimate
n <- c(1008,1116,146)  # number of captures
m <- c(0, 33,6)  # number of recaptures - not sure why this was 33 (instead of 24) changed back to 24. check with adam about where 33 came from
R <- c(1008, 1083, 0)  # of marked fish returned to the population
M <- c(0, cumsum(R)[-3])
seymour15 <- data.frame(n = n, m = m, R = R, M = M)

nM <- n * M
m.s <- sum(m)
lambda <- sum(nM)
N.schnabel <- lambda/(m.s + 1)
z <- 1.96
ci.sch <- c(lambda * (2 * m.s + z^2 - z * sqrt(4 * m.s + z^2))/(2 * m.s^2), 
            lambda * (2 * m.s + z^2 + z * sqrt(4 * m.s + z^2))/(2 * m.s^2))
#Schnabel Results

mS.table = data.table(parameter = c("N''", "ci low", "ci  up"), value = format(c(N.schnabel, 
                                                                                  ci.sch[1], ci.sch[2]), scientific = FALSE, digits = 3))

data_sum2 %>% 
  filter(area == "Seymour" & year == 2015) -> Seymour_15
Seymour_15 %>% 
  dplyr::select(year, event, area, M, n, m, CSA_legalcrab, legal_wt) %>% 
  mutate(Schnabel = as.numeric(mS.table$value[1]), 
         upper = as.numeric(mS.table$value[3]), 
         lower = as.numeric(mS.table$value[2])) %>% 
  mutate(Schnabel_lb = Schnabel*legal_wt, upper_lb = upper*legal_wt, lower_lb = lower*legal_wt, 
         adj = Schnabel_lb/CSA_legalcrab) -> Schnabel1
Schnabel1 %>% 
  filter(event == 1) %>% 
  dplyr::select(year, area, M, n, m, Schnabel, upper, lower) -> table2b

panderOptions("digits", 4)
pander(table2b, row.names = F )

```


*Footnote: M = total number of crab marked in the first event, n = total number of crab caught during recapture event, m = number of crab recaptured (ones with marks). * 



**Table 3: Summary of mark recapture and CSA population estimates**
```{r table 3, echo=F}
data_sum_Chaponly %>% 
  dplyr::select(year, area, CSA_legalcrab,Chap_lb, lower_lb, upper_lb) %>% 
  mutate(Estimator = "Chapman") %>% 
  rename(CSA_legal = CSA_legalcrab, 
         MR_legal = Chap_lb, 
         LCI = lower_lb, 
         UCI = upper_lb) ->table3_chap

Schnabel1 %>% 
  filter(event ==1) %>% 
  dplyr::select(year, area, CSA_legalcrab,Schnabel_lb, lower_lb, upper_lb) %>% 
  mutate(Estimator = "Schnabel") %>% 
  rename(CSA_legal = CSA_legalcrab, 
         MR_legal = Schnabel_lb, 
         LCI = lower_lb, 
         UCI = upper_lb) ->table3_schn

table3_chap %>% 
  bind_rows(table3_schn) -> table3

panderOptions("digits", 4)

pander(table3, row.names = F)
```

*Footnote: All values in Table 3 and 4 are in pounds (lb)*




**Table 4: Adjustment between population size estimates for each year**
```{r pander table 4, echo=F}
# from table 3 above just need to add adj....
table3 %>% 
  mutate(ADJ = MR_legal/CSA_legal, lci_ADJ = LCI/CSA_legal, 
         uci_ADJ = UCI/CSA_legal) %>% # add adjustment and LCI and UCI
  dplyr::select(year, area, CSA_legal, MR_legal, ADJ, lci_ADJ, uci_ADJ) %>% 
  arrange(area, year) ->table4
panderOptions("digits", 4)

pander(table4, row.names = F)
```



**Table 5: Adjustment used in the stock assessment, weighted mean for areas with more than one experiment**
```{r table 5, echo=F}
table4 %>% 
  group_by(area) %>% 
  summarise(weighted_ADJ = weighted.mean(ADJ, CSA_legal)) ->table5

panderOptions("digits", 4)
pander(table5, row.names = F)
```



# DISCUSSION

## Assumptions and error
Mark-recapture experiments can be bias high if the same population is not sampled in the mark and subsequent recapture events. Due to the timing of the mark recapture experiment, late fall to winter, some crab may be moving out of the sampling area and migrating to shallower waters. Each survey area is considered to be a closed population, and for the most part genetic studies have supported this. However, crab may move around in each area/bay enough that they are not suseptible to the sampling methods during both "events". Movement of crab out of the sampling area would decrease the number of tags recaptures and therefore bias the population estimates high. 
Bias is also introduced into the estimation process if only a few (< 10) recaptures occur. Estimates of abundance may fail to give even the order of magnitude of the population correctly if this occurs (Seber 1982).The Seymour experiment in 2015 produced only 6 recaptures if it was treated as a two-event experiment (one mark and one recapture event). However, the timing of the mark events, spaced about two weeks apart, allowed for this experiment to be estimating using a multiple event estimator (Schnabel instead of Chapman estimator), and therefore reduced the bias associated with this estimate.


## Adjustment factor assumptions
Computing an adjustment factor between the CSA (survey) estimates of biomass and those from the mark recapture experiments carries its own assumptions. First, it is assumed that the CSA and mark recapture estimates of biomass will always be related to each other in a similar way over every year. Replicating the experiments in each area would aid in supporting this assumption, as seen in Pybus, Seymour, and Gambier. However, results from Excursion show statistically different adjustment factor estimates in the two replicates, even though the CSA estimates of legal biomass is similar in both experiment years. This suggests that other variables need to be considered in assuming that the adjustment factor can be applied to all years similarily. 